Higher-Order unification for free!

### 2024-04-10 paper/main.tex
fix

\section{Ground work for the compiler}
\label{sec:grounwork}

@@

Unification variables are identified by a
natural number, that represents a memory addresses
The memory and its associated operations are described below:

@@

Since each occurrence of a \Ho unification variables has a scope,
its solution needs to be abstracted over it to enable the
instantiation of a single assignment to different scopes.
This is expressed by the \elpiIn{inctx} container, and in particular
its \elpiIn{abs} binding constructor.
On the contrary a solution to a \Fo variable is a plain term.

@@

\noindent
We call \elpiIn{fsubst} the memory of \Fo{}, while we call \elpiIn{subst}
the one of \Ho.
The compiler establishes a mapping between variables of the two languages.

@@

\begin{invariant}[Unification-variable arity]
  Each variable \elpiIn{A}
  in \Ho has a (unique) arity \elpiIn{N} and each occurrence~
  \elpiIn{(uva A L)} is such that \elpiIn{(len L N)} holds
  \label{inv:uvaarity}
\end{invariant}

@@

@@

\input{code/comp_links}

@@

% \input{code/beta_fo}

% \noindent
% The \elpiIn{name} predicate holds only on nominal constants (i.e. bound
% variables).\footnote{Elpi provides it as a builtin, but one could implement it by
% systematically loading the hypothetical rule \elpiIn{name x} every time
% a nominal constant is postulated via \elpiInFN{pi x\ }}
% The choice of using n-ary application, rather than binary, is to make it
% easy to access the application's head. The price we pay is that substituting
% an application in the head of an application should be amended by
% ``flattening'' \elpiIn{fapp} nodes, that is the job of \elpiIn{napp}.
% \footnote{Note that \elpiIn{napp} is an artefact of formalization of \Fo{}
% we do in this presentation and, as we explain later,
% no equivalent of \elpiIn{napp} is needed in \Ho.}
% Finally note that the cut operator is inessential, it could be
% removed at the cost of a verbose test on the head of \elpiIn{L}
% in the second rule about \elpiIn{fapp}: \elpiIn{L}'s head
% can be \elpiIn{fcon}, \elpiIn{flam} or a name.

Applying the substitution corresponds to dereferencing a term with respect to
the memory. To ease the comparison we split \Fo{} dereferencing into a
\elpiIn{fder} step and a \elpiIn{napp} one. The former step replaces references
to memory cells that are set with their values, and has a corresponding
operation in \Ho, namely \elpiIn{deref}. On the contrary \elpiIn{napp}
has no corresponding operation in \Ho, and only ensures that
terms of the form <<\elpiIn{fapp[fapp L1|L2]}>> are replaced by
<<\elpiIn{fapp L3}>> where \elpiIn{L3} is the concatenation of \elpiIn{L1}
and \elpiIn{L2}. The reasons for this asymmetry is
that an \elpiIn{fapp} node with a flexible head is always mapped
to a \elpiIn{uva} (as per \cref{sec:simulation,sec:beta}),
preventing nested applications to materialize.

\input{code/fderef}

\noindent
Applying the substitution in \Ho{} is very similar, with
the caveat that assignments have to be moved to the
current scope, i.e. renaming the \elpiIn{abs}-bound variables
with the names in the scope of the unification variable occurrence.

\noindent
Note that move strongly relies on invariant \ref{inv:uvaarity}: the length
of the arguments of all occurrences of a unification variable and the
number of abstractions in its assignment have to match. In turn
this grants that \elpiIn{move} never fails.

\subsection{Notational conventions}

\noindent
When variables $x$ and $y$ can occur in term $t$ we shall write
$t_{xy}$ to stress this fact.

We write $\sigma = \{~ A_{xy} \mapsto y ~\}$ for the assignment
\elpiIn{abs x\abs y\y} and $\sigma = \{~ A \mapsto \lambda x.\lambda y.y ~\}$
for \elpiIn{lam x\lam y\y}.

@@

\todo{notation for problem compilation}

@@

\section[Compilation: fo\_tm to tm]{Basic simulation of \Fo{} in \Ho{}}
\label{sec:simulation}

In this section we describe a basic compilation scheme that we refine
later, in the following sections. This scheme is sufficient to implement
a \hstep that respects $\beta$-conversion for terms in \llambda.
The extension to $\eta\beta$-conversion is described in \cref{sec:eta} and
the support for terms outside \llambda in \cref{sec:beta}.

@@

\subsection{Limitations of by this basic scheme}
\label{sec:basic-comp-limitations}
The basic compilation scheme is not about to
deal wit the following problem:
\printAlll
  {{{\lambda x y.X \appsep y \appsep x, \lambda x y.x},{\lambda x. f \appsep (X \appsep x) \appsep x,Y}}}
  {{}}
  {{}}
  {{}}
% \begin{gather}
% \lambda x y.F \appsep y \appsep x = \lambda x y.x \label{eq:unif-eta1}\\
% \lambda x. f \appsep (F \appsep x) \appsep x = G \label{eq:unif-eta2}
% % \lambda x. f \appsep (F \appsep x) \appsep x = f \appsep (\lambda y.y) \label{eq:unif-eta2}
% \end{gather}

\noindent Note that here $X$ is used with different arities, moreover
in the second problem the left hand side happens to be an
eta expansion (of $f (\lambda y.y)$) only after we discover (at run time)
that $X = \lambda x\lambda y.y$ (i.e. that $X$ discards the $x$ argument).
Both problems are addressed in the next two sections.

>>>>

\section{Basic compilation and simulation}
\label{sec:simulation}

\subsection{Memory (\mapStore) and substitution ($\rho$ and $\sigma$)}
\label{sec:grounwork}

@@

Unification variables are identified by a
natural number that represents a memory addresses.
The memory and its associated operations are described below:

@@

Since each occurrence of a \Ho unification variables has a scope,
its assignment needs to be abstracted over it to enable the
instantiation of the same assignment to different scopes.
This is expressed by the \elpiIn{inctx} container, and in particular
its \elpiIn{abs} binding constructor.
On the contrary a solution to a \Fo variable is a plain term.

@@

\noindent
We call \elpiIn{fsubst} the memory of \Fo{}, while we call \elpiIn{subst}
the one of \Ho.

The compiler establishes a mapping between variables of the two languages.

@@

\begin{invariant}[Unification-variable arity]\label{inv:uvaarity}
  Each variable \elpiIn{A}
  in \Ho has a (unique) arity \elpiIn{N} and each occurrence
  \elpiIn{(uva A L)} is such that \elpiIn{L} has length \elpiIn{N}.
\end{invariant}

@@

Applying the substitution corresponds to dereferencing a term with respect to
the memory.
%  To ease the comparison we split \Fo{} dereferencing into a
% \elpiIn{fder} step and a \elpiIn{napp} one. The former step replaces references
% to memory cells that are set with their values, and has a corresponding
% operation in \Ho, namely \elpiIn{deref}. On the contrary \elpiIn{napp}
% has no corresponding operation in \Ho, and only ensures that
% terms of the form <<\elpiIn{fapp[fapp L1|L2]}>> are replaced by
% <<\elpiIn{fapp L3}>> where \elpiIn{L3} is the concatenation of \elpiIn{L1}
% and \elpiIn{L2}. The reasons for this asymmetry is
% that an \elpiIn{fapp} node with a flexible head is always mapped
% to a \elpiIn{uva} (as per \cref{sec:simulation,sec:beta}),
% preventing nested applications to materialize.
% \input{code/fderef}
It is worth looking at the code fpr \Ho{} to remark how assignments are moved
to the current scope, i.e. renaming the \elpiIn{abs}-bound variables
with the names in the scope of the unification variable occurrence.

\noindent
Note that move strongly relies on invariant \ref{inv:uvaarity}: the length
of the arguments of all occurrences of a unification are the same. Hence
they have the same type in the meta-level and the number of \elpiIn{abs}
nodes in the assignment matches that length. In turn this grants that
\elpiIn{move} never fails.\todo{convoluted}

We write $\sigma = \{~ A_{xy} \mapsto y ~\}$ for the assignment
<<\elpiIn{abs x\abs y\y}>> and $\sigma = \{~ A \mapsto \lambda x.\lambda y.y ~\}$
for <<\elpiIn{lam x\lam y\y}>>.

\subsection{Links (\linkStore)}

@@

\input{code/comp_links}
\todo{rename link beta to link llam?}

@@

@@

\subsection{Notational conventions}\todo{move away}

When variables $x$ and $y$ can occur in term $t$ we shall write
$t_{xy}$ to stress this fact. 
\todo{add notation for problem compilation}

@@

% \section[Compilation: fo\_tm to tm]{Basic simulation of \Fo{} in \Ho{}}

% In this section we describe a basic compilation scheme that we refine
% later, in the following sections. This scheme is sufficient to implement
% a \hstep that respects $\beta$-conversion for terms in \llambda.
% The extension to $\eta\beta$-conversion is described in \cref{sec:eta} and
% the support for terms outside \llambda in \cref{sec:beta}.

@@

% \subsection{Limitations of by this basic scheme}
% \label{sec:basic-comp-limitations}
% The basic compilation scheme is not about to
% deal wit the following problem:
% \printAlll
%   {{{\lambda x y.X \appsep y \appsep x, \lambda x y.x},{\lambda x. f \appsep (X \appsep x) \appsep x,Y}}}
%   {{}}
%   {{}}
%   {{}}
% % \begin{gather}
% % \lambda x y.F \appsep y \appsep x = \lambda x y.x \label{eq:unif-eta1}\\
% % \lambda x. f \appsep (F \appsep x) \appsep x = G \label{eq:unif-eta2}
% % % \lambda x. f \appsep (F \appsep x) \appsep x = f \appsep (\lambda y.y) \label{eq:unif-eta2}
% % \end{gather}

% \noindent Note that here $X$ is used with different arities, moreover
% in the second problem the left hand side happens to be an
% eta expansion (of $f (\lambda y.y)$) only after we discover (at run time)
% that $X = \lambda x\lambda y.y$ (i.e. that $X$ discards the $x$ argument).
% Both problems are addressed in the next two sections.

\section{Handling of \maybebeta}\label{sec:llam}

\subsection{Detection}
\subsection{Compilation and decompilation}
\subsection{Progress}
